---
title: 'MDstatsDIAMS: Real Data Exploration Using Spectronaut Report'
author: Namgil Lee^[Kangwon National University, and Bionsight Inc., namgil.lee@kangwon.ac.kr],
  Hojin Yoo^[Bionsight Inc.], Juhyoung Kim^[Kangwon National University], Heejung
  Yang^[Kangwon National University, and Bionsight Inc., heejyang@kangwon.ac.kr]
date: "June 14, 2025"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
---

This vignette explores and visualizes real DIA-MS data given in a
Spectronaut report. This vignette reproduces figures and tables in Sections S1 and S2 of the Supplementary Material of the paper "*A Shrinkage-based Statistical Method for Testing Group Mean Differences in Quantitative Bottom-up Proteomics*" written by the authors.

## Preparation 

1. Set parameters to reduce analysis time

Because the real data size can be too large, users can choose
parameters to reduce analysis time in this vignette.

```{r set_analysis_scale}
# n_protein:
#   Size of a subset of proteins randomly selected.
#   If -1 or Inf, include all proteins. Default to 100.
n_protein <- 100

# remove_intermediate_reports:
#   If TRUE, remove report from environment if not used in further steps. 
#   Default is TRUE.
remove_intermediate_reports <- TRUE
```

2. Load packages.

```{r load_package, message=FALSE}
library(dplyr)
```

3. Load Report Data

Load a Spectronaut report from a remote repository.

```{r read_report}
df_real <- arrow::read_parquet(
  paste0(
    "/Users/namgil/Documents/Projects/MDstatsDIAMS/data/",
    "lip_quant_staurosporine_hela_sn_report.parquet"
  )
)
```

4. Filter the report and compute log10-transformed precursor quantity.

```{r filter_report}
df_filtered <- df_real %>% 
  filter(
    F.ExcludedFromQuantification == 'False'
    & EG.Qvalue < 0.01
    & F.NormalizedPeakArea > 1
  ) %>%
  mutate(F.Log10NormalizedPeakArea = log10(F.NormalizedPeakArea)) %>%
  select(
    R.Condition, R.Replicate, PG.ProteinGroups, PG.ProteinNames,
    EG.ModifiedSequence, FG.Charge, F.FrgIon, F.FrgLossType, F.Charge,
    EG.Qvalue, F.Log10NormalizedPeakArea
  )

print(dim(df_filtered))
```

5. Set the order of the condition values so that "DMSO" comes before the other
conditions.

```{r sort_conditions}
conditions = unique(df_filtered$R.Condition)

df_filtered <- df_filtered %>%
  mutate(
    R.Condition = factor(
      R.Condition, 
      labels = c("DMSO", conditions[-which(conditions == "DMSO")])
    )
  )

print(levels(df_filtered$R.Condition))
```

6. Randomly select a subset of the predefined number of proteins to reduce analysis time.

```{r choose_report_subset}
if (
  n_protein < 0
  || is.infinite(n_protein) 
  || n_protein >= n_distinct(df_filtered$PG.ProteinGroups)
) {
  df_subset = df_filtered
} else {
  set.seed(111)
  protein_subset = sample(unique(df_filtered$PG.ProteinGroups), n_protein)
  df_subset = df_filtered %>% filter(PG.ProteinGroups %in% protein_subset)
}

print(dim(df_subset))
```

7. Remove the large report data from the R environment that will not be used in the next steps.

```{r remove_intermediate, echo=FALSE}
if (isTRUE(remove_intermediate_reports)) {
  rm(df_real, df_filtered)
}
```

## Summary Statistics

Print summary statistics by condition.

```{r print_summary_by_condition, echo=FALSE, message=FALSE}
# Data counts
all_counts = df_subset %>% 
  group_by(R.Condition) %>%
  summarise(
    num_protein = n_distinct(PG.ProteinGroups),
    num_precursor = n_distinct(PG.ProteinGroups, EG.ModifiedSequence, FG.Charge),
    num_fragment_ion = n_distinct(PG.ProteinGroups, EG.ModifiedSequence, FG.Charge,
                                  F.FrgIon, F.FrgLossType, F.Charge))

print(all_counts)

# Mean and std of log10 precursor quantity
all_peptide_quantity = df_subset %>%
  group_by(R.Condition, PG.ProteinGroups, PG.ProteinNames, EG.ModifiedSequence,
           FG.Charge, R.Replicate) %>%
  summarise(precursor_quantity = sum(10 ** F.Log10NormalizedPeakArea)) %>%
  group_by(R.Condition) %>%
  summarise(
    precursor_mean_quantity = mean(log10(precursor_quantity), na.rm = TRUE), 
    precursor_sd_quantity = sd(log10(precursor_quantity), na.rm = TRUE)
  )

print(all_peptide_quantity)
```

## Precursor Quantity Distribution

1. Precursor quantity distribution at fixed condition

Distribution of precursor quantity: all precursors measured in four replicates under fixed condition.

```{r precursor_quantity, echo=FALSE, message=FALSE}
con = "DMSO"

precursor_quantity = df_subset %>%
  filter(R.Condition==con) %>%
  group_by(
    PG.ProteinGroups, PG.ProteinNames, EG.ModifiedSequence, FG.Charge,
    R.Condition, R.Replicate
  ) %>%
  summarise(precursor_quantity = sum(10 ** F.Log10NormalizedPeakArea))
```

```{r precursor_quantity_hist, echo=FALSE, message=FALSE}
lquant = log10(precursor_quantity$precursor_quantity)
n_lquant = length(lquant)
par(mfrow = c(1, 2))

# Histogram
par(mar = c(5.6, 4.6, 4.1, 2.1))
resolution = 0.1
histout = hist(lquant, seq(0, 10, resolution), 
               ylab = "Frequency", xlab = "Peptide Quantity, log10",
               cex.lab = 1.2, main = paste("Condition:", con))
mn_est <- histout$mids[order(histout$counts, decreasing = TRUE)[1]]
sd_est <- IQR(lquant) / (qnorm(0.75) - qnorm(0.25))
lines(seq(0, 10, 0.05), 
      n_lquant * resolution * dnorm(seq(0, 10, 0.05), mn_est, sd_est), 
      lwd = 2, col = "red", lty = 2)
legend("topright", legend=c("Empirical Density", "Normal Density"), 
       pch = c(0, NA), lty = c(0, 2), col = c("black", "red"), 
       lwd = c(1,2), cex = 0.6)

# QQ-plot
par(mar = c(5.6, 4.6, 4.1, 2.1))
set.seed(111)
qqnorm_out = qqnorm(sample(lquant, round(n_lquant/5)), cex.lab = 1.2,
     main = paste("Condition:", con), cex = 0.5)
qqline(lquant)
legend("topleft",
       legend=paste("Pearson r = ", round(cor(qqnorm_out$x, qqnorm_out$y), 3)),
       cex = 0.6)
```

2. An Example of Dose-Response Plot for A Selected Precursor

Distribution of precursor quantity: a selected precursor across all
conditions.

```{r selected_precursor_distribution, echo=FALSE, message=FALSE}
pname = df_subset$PG.ProteinNames[1]
mod = df_subset$EG.ModifiedSequence[1]
chg = df_subset$FG.Charge[1]

selected_prec_quant_df = df_subset %>%
  filter(PG.ProteinNames == pname
         & EG.ModifiedSequence == mod
         & FG.Charge == chg) %>%
  group_by(
    PG.ProteinGroups, PG.ProteinNames,
    EG.ModifiedSequence, FG.Charge, R.Condition, R.Replicate
  ) %>%
  summarise(precursor_quantity = sum(10 ** F.Log10NormalizedPeakArea))

## Scatter Plot
par(mar = c(5.6, 4.6, 4.1, 2.1))
lquant = log10(selected_prec_quant_df$precursor_quantity)
ymin = floor(min(lquant))
ymax = floor(max(lquant + 0.99))
plot(log10(precursor_quantity) ~ as.numeric(R.Condition), 
     data = selected_prec_quant_df, 
     type = "p", xlab = "Condition", ylab = "Peptide Quantity, log10", 
     ylim = c(ymin, ymax), main = paste0(pname, ":", mod, ".", chg), 
     cex.lab=1.5, cex.main=1.5, cex=0.7, pch = 16, xaxt = "n",
     col = rgb(red = 1, green = 0, blue = 0, alpha = 0.8))
axis(side = 1, 
     at = 1:length(unique(selected_prec_quant_df$R.Condition)), 
     labels = unique(selected_prec_quant_df$R.Condition)) 
mn_est <- mean(lquant, na.rm = TRUE)
sd_est <- sd(lquant, na.rm = TRUE)
abline(h = mn_est, lwd = 1.5, col = "blue", lty = 2)
abline(h = mn_est + sd_est, lwd = 1.5, col = "blue", lty = 2)
abline(h = mn_est - sd_est, lwd = 1.5, col = "blue", lty = 2)
grid()
```

## Sampling Distribution for Hierarchical Model

Distribution of precursor mean and deviance:

* precursor mean: all precursors averaged over replicates under fixed condition.
* precursor deviance: all precursors subtracted by the mean.

```{r precursor_mean_distribution, echo=FALSE}
precursor_deviation_df = precursor_quantity %>% 
  group_by(PG.ProteinGroups, PG.ProteinNames, EG.ModifiedSequence, FG.Charge) %>%
  mutate(precursor_mean_quantity = mean(log10(precursor_quantity), na.rm=TRUE)) %>% 
  mutate(peptide_quantity_deviance = log10(precursor_quantity) - precursor_mean_quantity)

lquant_mean <- precursor_deviation_df %>% 
  distinct(PG.ProteinNames, EG.ModifiedSequence, FG.Charge, .keep_all = TRUE) %>%
  pull(precursor_mean_quantity)

lquant_deviation <- precursor_deviation_df %>%
  pull(peptide_quantity_deviance)

## 1. Distribution of mean of precursor quantity
par(mar = c(5.6, 4.6, 4.1, 2.1))
resolution = 0.1
histout = hist(lquant_mean, 
     seq(0, 10, resolution), 
     ylab = "Frequency", 
     xlab = "Mean Peptide Quantity, log10", 
     cex.lab = 1.7,
     main = paste("Condition:", con))
mn_est = histout$mids[order(histout$counts, decreasing = TRUE)[1]]
sd_est = IQR(lquant_mean) / (qnorm(0.75) - qnorm(0.25))
lines(seq(0, 10, 0.05), 
      length(lquant_mean) * resolution * dnorm(seq(0, 10, 0.05), mn_est, sd_est), 
      lwd = 2, col = "red", lty = 2)
legend("topright", legend=c("Empirical Density", "Normal Density"), 
       pch = c(0, NA), lty = c(0, 2), col = c("black", "red"), 
       lwd = c(1,2))

## 2. Distribution of deviation of precursor quantity
par(mar = c(5.6, 4.6, 4.1, 2.1))
resolution = 0.02
histout = hist(lquant_deviation, 
     seq(-4.2, 4.2, resolution), 
     xlim = c(-1, 1), 
     ylab = "Frequency", 
     xlab = "Deviance of Peptide Quantity, log10", 
     cex.lab = 1.7,
     main = paste("Condition:", con))
mn_est = median(lquant_deviation)
sd_est = IQR(lquant_deviation) / (qnorm(0.75) - qnorm(0.25))
lines(seq(-4.2, 4.2, 0.01), 
      length(lquant_deviation) * resolution * dnorm(seq(-4.2, 4.2, 0.01), mn_est, sd_est), 
      lwd = 2, col = "red", lty = 2)
legend("topright", legend=c("Empirical Density", "Normal Density"), 
       pch = c(0, NA), lty = c(0, 2), col = c("black", "red"), 
       lwd = c(1,2))
```

## Distribution of Ionization Efficiency

1. Distribution of the number of fragment ions of each precursor

* The number of fragment ions of each precursor is computed. The precursors are
separated by their charges to inspect distribution specifically.

```{r fragment_ion_count_by_charge, echo=FALSE, message=FALSE}
con <- "DMSO"
max_charge <- max(df_subset$FG.Charge)
  
par(mfrow = c(2,3))
for (charge in 1:max_charge) {
  num_ions_prec <- df_subset %>%
    filter(R.Condition == con & FG.Charge==charge) %>%
    group_by(R.Replicate, PG.ProteinGroups, EG.ModifiedSequence) %>%
    summarise(num_ions = n_distinct(F.FrgIon, F.FrgLossType, F.Charge))
  
  barplot(table(num_ions_prec$num_ions), 
          xlab="Number of ions",
          ylab="Number of precursors",
          main=paste("Precursor Charge:", charge))
}
```

2. Calculate ionization efficiency distribution 

* A fragment ion quantity is calculated by its normalized peak area.
* The ionization efficiency of a fragment ion is the proportion of the fragment
ion quantity out of the sum of the fragment ion quantities in its corresponding
precursor.
* For analysis, the precursors are separated by their numbers of fragment ions
to inspect distribution specifically.

```{r ionization_efficiency_distribution, echo=FALSE}
con = "DMSO"

# Compute the number of fragment ions of each precursor, and 
# ionization efficiency of each fragment ion
num_ions_prec <- df_subset %>%
  filter(R.Condition == con) %>%
  group_by(R.Replicate, PG.ProteinGroups, EG.ModifiedSequence, FG.Charge) %>%
  mutate(
    num_ions = n_distinct(F.FrgIon, F.FrgLossType, F.Charge), 
    ion_eff = (10 ** F.Log10NormalizedPeakArea) / sum(10 ** F.Log10NormalizedPeakArea)
  )

# Plot distribution of ionization efficiency
par(mfrow = c(2,3))
for (ni in 1:6) {
  ylab= ifelse (ni %in% c(1,4), "Frequency" , "")
  xlab = ifelse (ni %in% c(4,5,6), "Ionization Efficiency" , "")
  
  ion_eff = num_ions_prec %>% filter(num_ions==ni) %>% pull(ion_eff)
  par(mar = c(5.6, 4.6, 4.1, 2.1))
  resolution = 0.025
  histout <- hist(ion_eff, breaks = seq(0, 1, resolution), right = FALSE,
                 main = paste0("Precursors num ions: ", ni), cex.main = 1.2, 
                 xlab = xlab, ylab = ylab, cex.lab = 1.5, xlim = c(0, 1))
  
  if (ni > 1) {
    params <- Rfast::beta.mle(ion_eff)
    alpha_est <- params$param["alpha"]
    beta_est <- params$param["beta"]
    
  } else {
    alpha_est <- 1
    beta_est <- 0
    legend("topleft", legend=c("Empirical Density", "Beta Density"), 
           pch = c(0, NA), lty = c(0, 2), col = c("black", "red"),
           lwd = c(1, 1.5), cex=1.1)
  }
  
  mn_est <- alpha_est / (alpha_est + beta_est)
  if (mn_est == 1) {
    abline(v=1, lwd=2, col="red", lty=2)
  } else {
    xs <- seq(0, 1, 0.01)
    ys <- length(ion_eff) * resolution * dbeta(xs, alpha_est, beta_est)
    lines(xs, ys, lwd = 2, col = "red", lty = 2)
  }
}
```

3. Fit Beta distribution to an ionization efficiency distribution

* A beta distribution has two shape parameters, alpha and beta. The mean of a
Beta distribution is equal to alpha / (alpha + beta).
* In this analysis, 30 proteins are randomly selected among the proteins having
at least 30 peptides. For each protein, alpha and beta are estimated
by ionization efficiency distribution of its fragment ions.

```{r fit_beta_prot30, echo=FALSE}
con = "DMSO"

# Select 30 proteins
peptide_count_per_protein <- df_subset %>%
  filter(R.Condition==con) %>%
  group_by(PG.ProteinNames) %>%
  summarise(peptide_count = length(unique(EG.ModifiedSequence)))
set.seed(100)
proteins30 = peptide_count_per_protein %>% filter(peptide_count >= 30) %>% pull(PG.ProteinNames)
proteins_selected = sample(proteins30, min(30, length(proteins30)))

# Estimate parameters (alpha, beta) for each protein
estimates = matrix(0, 0, 4)
for (prot in proteins_selected) {
  num_ions_prec_prot <- df_subset %>%
    filter(R.Condition == con & PG.ProteinNames == prot) %>%
    group_by(R.Replicate, PG.ProteinGroups, EG.ModifiedSequence, FG.Charge) %>%
    mutate(
      num_ions = nrow(unique(cbind(F.FrgIon, F.FrgLossType, F.Charge))), 
      ion_eff = (10 ** F.Log10NormalizedPeakArea) / sum(10 ** F.Log10NormalizedPeakArea)
    )
  
  # Iterate across each number of fragment ions
  for (ni in 1:6) {
    ion_eff_fixedprot <- num_ions_prec_prot %>% filter(num_ions==ni) %>% pull(ion_eff)
      
    if (length(ion_eff_fixedprot) > 0) {
      if (ni > 1) {
        params = Rfast::beta.mle(ion_eff_fixedprot)
        alpha_est = params$param["alpha"]
        beta_est = params$param["beta"]
      } else {
        alpha_est = 1
        beta_est = 0
      }
      estimates <- rbind(estimates, c(
        protein_accessions = prot, num_ions = ni,
        alpha = alpha_est, beta = beta_est
      ))
    }
  }
}

estimates = data.frame(estimates)
estimates$num_ions = as.numeric(estimates$num_ions)
estimates$alpha = as.numeric(estimates$alpha)
estimates$beta = as.numeric(estimates$beta)

max_lim = min(c(max(estimates$alpha), max(estimates$beta)))

par(mar = c(5.0, 5.0, 4.1, 2.1))
with(estimates[estimates$num_ions==2,], 
     plot(alpha, beta, 
          xlab=expression(paste(estimated, " ", alpha)), 
          ylab=expression(paste(estimated, " ", beta)), 
          pch=1, col=1, cex.lab=2.0, xlim=c(0, max_lim), ylim=c(0, max_lim), 
          main="Estimated Beta shape parameters for ionization efficiency"))
abline(a=0, b=1/1, col=2, lty=2, lwd=1.5)
for (ni in 3:6) {
  with(estimates[estimates$num_ions==ni,], 
       points(alpha, beta, 
              xlab=expression(paste(estimated, " ", alpha)), 
              ylab=expression(paste(estimated, " ", beta)), 
              pch=ni-1, col=1))
  abline(a=0, b=(ni-1)/1, col=2, lty=2, lwd=1.5)
}
legend("topright", legend=c(2,3,4,5,6), lty=2, pch=1:5, lwd=1.5, col=1,
       title = "Num fragment ions")
```

* In the figure, each dot is a protein, and there are 30 points on each straight
line.
* The dashed lines are $y=x,\, y=2x,\, \ldots,\, y=5x$. 
* The figure shows that the estimated (alpha, beta) values agree with
the dashed lines.


4. Correlation of Ionization Efficiency Between Conditions

* Investigate correlation between fragment ion quantities of two conditions with
the same precursors, fragment ions, and replicate

```{r correlation_ionization, echo=FALSE}
con1 = "DMSO"
con2 = "100pM"

num_ions_prec1 <- df_subset %>%
  filter(R.Condition == con1) %>%
  group_by(R.Replicate, PG.ProteinGroups, EG.ModifiedSequence, FG.Charge) %>%
  mutate(
    num_ions1 = nrow(unique(cbind(F.FrgIon, F.FrgLossType, F.Charge))), 
    ion_eff1 = (10 ** F.Log10NormalizedPeakArea) / sum(10 ** F.Log10NormalizedPeakArea)
  )

num_ions_prec2 <- df_subset %>%
  filter(R.Condition == con2) %>%
  group_by(R.Replicate, PG.ProteinGroups, EG.ModifiedSequence, FG.Charge) %>%
  mutate(
    num_ions2 = nrow(unique(cbind(F.FrgIon, F.FrgLossType, F.Charge))), 
    ion_eff2 = (10 ** F.Log10NormalizedPeakArea) / sum(10 ** F.Log10NormalizedPeakArea)
  )

num_ions_prec_merged12 <- num_ions_prec1 %>% 
  inner_join(num_ions_prec2, 
             by = c("R.Replicate", "PG.ProteinGroups", "EG.ModifiedSequence", 
                    "FG.Charge", "F.FrgIon", "F.FrgLossType", "F.Charge"))

set.seed(100)
par(mfrow = c(2,3), mar=c(5.1, 4.8, 4.1, 2.1))
for (ni in 1:6) {
  xlab <- ifelse(ni %in% c(4, 5, 6), paste("Ionization Efficiency,", con1), "")
  ylab <- ifelse(ni %in% c(1, 4), paste("Ionization Efficiency,", con2), "")

  subdat <- num_ions_prec_merged12 %>% filter(num_ions1==ni &num_ions2==ni)
  sample_id <- sample(nrow(subdat), floor(nrow(subdat) * 0.05))
  r <- ifelse(ni == 1, 1, round(cor(subdat$ion_eff1, subdat$ion_eff2), 2))
  plot(subdat$ion_eff1[sample_id], subdat$ion_eff2[sample_id], 
       xlab=xlab, ylab=ylab, xlim=c(0,1), ylim=c(0,1), cex=0.4, cex.lab=1.4,
       main=paste0("Precursor num ions: ", ni, ", r=", r))
}
```
