---
title: "MDstatsDIAMS: Real Data Comparison of Methods for Testing Mean Differences"
author:
  - Namgil Lee^[Kangwon National University, and Bionsight Inc., namgil.lee@kangwon.ac.kr], Hojin Yoo^[Bionsight Inc.], Juhyoung Kim^[Kangwon National University], Heejung Yang^[Kangwon National University, and Bionsight Inc., heejyang@kangwon.ac.kr]
date: "June 14, 2025"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
---

This vignette includes numerical comparison of the methods for testing mean differences using real DIA-MS data, which are given in either Spectronaut, MaxQuant, or Skyline report format. This vignette reproduces figures and tables in Section 3.2 Mass Spectrometry Data Analysis of the paper "*A Shrinkage-based Statistical Method for Testing Group Mean Differences in Quantitative Bottom-up Proteomics*" written by the authors.

## Preparation

1. Set parameters to reduce analysis time

Because the real data size can be too large for the vignette, users can choose
parameters to reduce analysis time.

```{r set_analysis_scale}
# n_protein:
#   Size of a subset of proteins randomly selected.
#   If -1 or Inf, include all proteins. Default to 100.
n_protein <- 100

# input_report_format:
#   User must choose one of "sn"(Spectronaut), "mq" (MaxQuant), or 
#   "sk" (Skyline).
input_report_format <- "sn"

# result_save_folder:
#   Path to the folder to save ttest results.
result_save_folder <- "/Users/namgil/Documents/Projects/MDstatsDIAMS/data/"

# remove_intermediate_reports:
#   If TRUE, remove report from environment if not used in further steps.
#   Its value does not influence the results.
#   Default is TRUE.
remove_intermediate_reports <- TRUE

# control_condition,treat_condition:
#   Comparison between control and treatment. Select conditions
#   from {"DMSO", "10nM", "100nM", "100mM"} if Spectronaut;
#   from {"CON1", "CON4", "CON5", "CON8"} if MaxQuant or Skyline
control_condition <- "DMSO"
treat_condition <- "100mM"

# subconditions:
#   List of four conditions for additional pairwise comparisons.
#   Set {"DMSO", "10nM", "100nM", "100mM"} if Spectronaut;
#   {"CON1", "CON4", "CON5", "CON8"} if MaxQuant or Skyline
subconditions <- c("DMSO", "10nM", "100nM", "100mM")
```

2. Load packages

```{r load_package, message=FALSE}
library(dplyr)
library(MDstatsDIAMS)
library(pROC)
```

3. Load target list

Load an on-target protein list.

```{r}
known_target_df <- read.csv(
  paste0(
    "https://zenodo.org/records/15653980/files/",
    "known_target_list_staurosporine_davis2011.csv"
  )
)
```

4. Load report data 

Load and preprocess report data, and convert it into standard format. Spectronaut, MaxQuant, and Skyline reports can be converted into standard format directly. 

* Reports are filtered by q-value < 0.01, fragment peak area > 1. 
* Decoy proteins are removed from MaxQuant and Skyline.
* Top-3 fragment ions are selected in MaxQuant.

```{r read_report}
if (input_report_format == "sn") {
  ## Load Spectronaut report
  df_real <- arrow::read_parquet(
    paste0(
      "https://zenodo.org/records/15653980/files/",
      "lip_quant_staurosporine_hela_sn_report.parquet"
    )
  )

  df_filtered <- convert_sn_to_standard(df_real, filter_identified = TRUE)

} else if (input_report_format == "mq") {
  ## Load MaxQuant report
  mq_root <- paste0("/Users/namgil/Documents/Projects/MDstatsDIAMS/data/",
                    "lip_quant_staurosporine_hela_mq_report_4conds/")

  mq_evidence_path <- paste0(mq_root, "evidence.txt")
  mq_msms_path <- paste0(mq_root, "msms.txt")

  df_real <- read.delim(mq_evidence_path)
  mq_msms <- read.delim(mq_msms_path)

  ##  It is required to generate `annotation.txt' file in prior for **MSstats**.
  mq_an <- create_annotation_df(
    evidence = df_real,
    n_replicates_per_condition = 4
  )
  df_filtered <- convert_mq_to_standard(df_real, mq_msms, mq_an)

} else if (input_report_format == "sk") {
  ## Load Skyline report
  df_real <- arrow::read_parquet(
    paste0("/Users/namgil/Documents/Projects/MDstatsDIAMS/data/",
           "lip_quant_staurosporine_hela_sk_transition_4conds.parquet")
  )

  # Make an annotation data frame
  sk_an <- data.frame(
    Condition = rep(paste0("CON", c(1, 4, 5, 8)), each = 4),
    Replicate = paste0("StauroDoseResp-", c(1:4, 17:24, 33:36)),
    Run = paste0("StauroDoseResp-", c(1:4, 17:24, 33:36))
  )

  df_filtered <- convert_sk_to_standard(df_real, annotation = sk_an)
}

print(dim(df_filtered))

```


5. Randomly select a subset of the predefined number of proteins to reduce
analysis time.

```{r choose_report_subset}
if (
  n_protein < 0
  || is.infinite(n_protein) 
  || n_protein >= n_distinct(df_filtered$protein_id)
) {
  df_subset <- df_filtered
} else {
  set.seed(111)
  protein_subset <- sample(unique(df_filtered$protein_id), n_protein)
  df_subset <- df_filtered %>% filter(protein_id %in% protein_subset)
}

print(dim(df_subset))
```

6. Remove the large report data from the R environment that will not be used in the next steps.

```{r remove_intermediate, echo=FALSE}
if (isTRUE(remove_intermediate_reports)) {
  rm(df_real, df_filtered)
}
```


## Comparison Between Control and Treatment

1. Run statistical methods for comparing the means between two conditions,
control (DMSO) and treatment (100 uM).

* Select a subset of the report for two conditions, control (DMSO) and treatment
  (100 uM)

```{r filter_two_conditions}
df_two_conds <- df_subset %>%
  filter(condition %in% c(control_condition, treat_condition))

print(paste("Comparing mean differences between two conditions:",
            control_condition,
            treat_condition))
```

* Run statistical methods.

```{r run_methods}
## Compute ttest result comparing two conditions ##
save_path <- paste0(
  result_save_folder,
  input_report_format, "_",
  "ttest_result_n", n_protein, "_", control_condition, "_", treat_condition,
  ".RData"
)

if (file.exists(save_path)) {
  load(save_path)
} else {
  # Run methods
  ttest_result <- run_ttests(
    df_two_conds,
    method_names = NULL,
    boot_denom_eps = 0.3,
    base_condition = control_condition
  )
  save(ttest_result, file = save_path)
}
```


2. Append local FDR score and on-target information

```{r compute_lfdr}
lfdr_result <- compute_lfdr_result(
  ttest_result,
  known_target_df = known_target_df,
  is_target_column = "is_target"
)
```


3. Compute sensitivity and specificity from the local fdr results

```{r sensitivity_result}
protein_sensitivity_result <- compute_sensitivity_result(
  lfdr_result,
  q_value_column = "lfdr",
  is_target_column = "is_target",
  group_column = c("experiment", "protein_id")
)
```

* Print sensitivity and specificity in a brief form

```{r print_sensitivity}
print(paste("Comparison:", control_condition, "/", treat_condition))

spec_levels <- seq(0.8, 0.2, -0.2)
sens_matrix <- matrix(
  NA, length(protein_sensitivity_result), length(spec_levels),
  dimnames = list(names(protein_sensitivity_result), spec_levels)
)

for (level in spec_levels) {
  for (method in names(protein_sensitivity_result)) {
    sensitivity_table <- protein_sensitivity_result[[method]][[1]]
    spec_id_at_level <- max(which(sensitivity_table$specificity > level))
    sens_at_level <- sensitivity_table$sensitivity[spec_id_at_level]
    sens_matrix[method, as.character(level)] <- sens_at_level
  }
}

print(round(sens_matrix, 2))
```

4. Print contingency table at fixed significance level

```{r contingency_table_lfdr, echo=FALSE}
print("---------- Significance level: 0.10 ----------")
tables10 <- compute_contingency_tables(lfdr_result, 0.10, "lfdr")
print(tables10)

print("---------- Significance level: 0.05 ----------")
tables05 <- compute_contingency_tables(lfdr_result, 0.05, "lfdr")
print(tables05)

print("---------- Significance level: 0.01 ----------")
tables01 <- compute_contingency_tables(lfdr_result, 0.01, "lfdr")
print(tables01)
```

## Testing Across Multiple Conditions

* Aggregate p-values of consecutive comparisons

```{r aggregate_pvalue, message=FALSE}
protein_df_agg <- list()
aucs <- c()

### Collect p-values
num_cond <- length(subconditions)
for (i in 1:(num_cond - 1)) {
  ## Compute p-value from consecutive conditions
  control_condition <- subconditions[i]
  treat_condition <- subconditions[i + 1]

  df_two_conds <- df_subset %>%
    filter(condition %in% c(control_condition, treat_condition))

  ## Compute ttest result comparing two conditions ##
  save_path <- paste0(
    result_save_folder,
    input_report_format, "_",
    "ttest_result_n", n_protein, "_", control_condition, "_", treat_condition,
    ".RData"
  )
  if (file.exists(save_path)) {
    load(save_path)
  } else {
    ttest_result <- run_ttests(
      df_two_conds,
      method_names = NULL,
      boot_denom_eps = 0.3,
      base_condition = control_condition
    )
    save(ttest_result, file = save_path)
  }
  ##

  ## Compute "signed" p-value in precursor level
  for (method in names(ttest_result)) {
    signed_pvalue_table <- ttest_result[[method]][[1]] %>% 
      select(experiment, protein_id, precursor_id, p.value, estimate) %>%
      mutate(comparison_1 = control_condition,
             comparison_2 = treat_condition,
             sign_of_test = sign(estimate)) %>%
      filter(!is.na(p.value))

    # protein_df_agg[[method]] columns: experiment, protein_id, precursor_id,
    #   p.value, estimate, comparison_1, comparison_2, sign_of_test
    protein_df_agg[[method]] <- rbind(
      protein_df_agg[[method]],
      signed_pvalue_table
    )
  }
}

# Aggregate all the comparisons for each precursor
for (method in names(protein_df_agg)) {
  protein_df_agg[[method]] <- protein_df_agg[[method]] %>%
    group_by(experiment, protein_id, precursor_id) %>%
    mutate(sign_of_minimal_pvalue = sign_of_test[which.min(p.value)]) %>%
    group_by(experiment, protein_id, precursor_id) %>%
    mutate(prod_recomputed_pvalue = prod(
      0.5 - sign_of_test * sign_of_minimal_pvalue * (0.5 - 0.5 * p.value),
      na.rm = TRUE)) %>%
    group_by(experiment, protein_id, precursor_id) %>%
    summarise(p.value = sum(prod_recomputed_pvalue * 2, na.rm = TRUE))
} ## experiment, protein_id, precursor_id, p.value

# Aggregate to protein level
for (method in names(protein_df_agg)) {
  protein_df_agg[[method]] <- protein_df_agg[[method]] %>%
    group_by(experiment, protein_id) %>%
    summarise(p.value = min(p.value, na.rm = TRUE))
}

# Append on-target information
for (method in names(protein_df_agg)) {
  protein_df_agg[[method]]$is_target <- (
    protein_df_agg[[method]]$protein_id %in% known_target_df$UniprotID
  )

  roc_prot <- pROC::roc(
    protein_df_agg[[method]]$is_target,
    1 - protein_df_agg[[method]]$p.value,
    auc = TRUE,
    plot = FALSE,
    quiet = TRUE
  )
  aucs[method] <- roc_prot$auc
}

print(paste(c("Conditions:", subconditions), collapse = " "))
print("AUCs:")
print(aucs)
```
