---
title: 'MDstatsDIAMS: Simulated Data Comparison of Methods for Testing Mean Differences'
author:
  - Namgil Lee^[Kangwon National University, and Bionsight Inc., namgil.lee@kangwon.ac.kr], Hojin Yoo^[Bionsight Inc.], Juhyoung Kim^[Kangwon National University], Heejung Yang^[Kangwon National University, and Bionsight Inc., heejyang@kangwon.ac.kr]
date: "June 30, 2025"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
---

This vignette includes numerical comparison of the methods for testing mean differences using simulated report data. This vignette reproduces figures and tables in Section 3.1.3 Numerical Evaluation of the paper "*A Shrinkage-based Statistical Method for Testing Group Mean Differences in Quantitative Bottom-up Proteomics*" written by the authors.

## Preparation

1. Set parameters

Simulated data sets are generated using various parameters, including mean difference, standard deviation for mean, and beta distribution.

```{r set_parameters}
# BETA_SUM:
#   **Choose one from: c(Inf, 100, 30, 10)**
#   Beta distribution parameters will be determined at
#   (alpha, beta) = (BETA_SUM * 0.1 + 1, BETA_SUM * 0.9 + 1).
#   If Inf, random samples takes constant values of 0.1.
BETA_SUM <- Inf

# DIRI_SUM:
#   **Choose one from: c(Inf, 300, 100, 30, 10)**
#   Dirichlet distribution parameters will be determined at
#   (alpha1, alpha2, alpha3) = rep(floor(DIRI_SUM / 3) + 1, 3).
#   If Inf, random samples takes constant values of 1/3.
DIRI_SUM <- 3

# S1:
#   **Choose one from: c(0.5, 1.0, 1.5)**
#   standard deviation for normal distribution for generating group mean values.
S1 <- 1.0

# ALL_S2:
#   vector of standard deviation for normal distribution for noise.
ALL_S2 <- 0.05 * (2 ** (0:4))

# ALL_N_REPLICATES:
#   vector of numbers of replicates
ALL_N_REPLICATES <- c(4, 8, 16, 32, 64)

# COR_FRAG:
#   correlation between fragment ion peak area from two conditions
COR_FRAG <- 0.89

# COMPARISON
#   **Choose one from: c("CON1/CON2", "CON1/CON3", "CON1/CON4")**
#   comparison of two conditions.
#   - "CON1/CON2": no mean difference between two groups
#   - "CON1/CON3": mean difference is -log10(2)
#   - "CON1/CON4": mean difference is -2*log10(2)
COMPARISON <- "CON1/CON4"

# SAVE_FOLDER:
SAVE_FOLDER <- "/Users/namgil/Documents/Projects/MDstatsDIAMS/vignettes/"
```

2. Load packages

```{r load_package, message=FALSE}
library(MDstatsDIAMS)
```

3. Run simulation

```{r run_simulation}
if (!dir.exists(SAVE_FOLDER)) {
  message(paste("Creating save folder", SAVE_FOLDER))
  dir.create(SAVE_FOLDER, recursive = TRUE)
}

NR_RESU <- vector("list", length(ALL_S2))
for (i_s2 in seq_along(ALL_S2)) {
  s2 <- ALL_S2[i_s2]

  all_nr_tables <- vector("list", length(ALL_N_REPLICATES))
  names(all_nr_tables) <- as.character(ALL_N_REPLICATES)
  for (nr in ALL_N_REPLICATES) {

    # Set parameters
    params <- default_params
    if (is.infinite(BETA_SUM)) {
      params$acquisition_beta_fnt <-
        function(n, shape1, shape2) rep(0.1, n)
      params$acquisition_beta_shape1 <- -1
      params$acquisition_beta_shape2 <- -1
    } else {
      params$acquisition_beta_fnt <- rbeta
      params$acquisition_beta_shape1 <- BETA_SUM * 0.1 + 1
      params$acquisition_beta_shape2 <- BETA_SUM * 0.9 + 1
    }

    if (is.infinite(DIRI_SUM)) {
      params$ionization_dirichlet_fnt <-
        function(n, alpha) matrix(1 / 3, n, 3)
      params$ionization_dirichlet_alpha <- c(-1, -1, -1)
    } else {
      params$ionization_dirichlet_fnt <- dirmult::rdirichlet
      params$ionization_dirichlet_alpha <-
        rep(floor(DIRI_SUM / 3), 3) + 1
    }

    params$n_experiment <- 500
    params$n_condition <- 2  # Changed from 5 to 2
    i_comp <- as.numeric(substr(COMPARISON, 9, 9))
    params$prec_mean_condition_shift <-
      c(0, i_comp - 2) * log10(2)  # length must be equal to n_condition
    params$n_replicate <- nr
    params$prec_mean_std <- S1
    params$noise_std <- s2

    params$ionization_cor_bet_condition <- COR_FRAG

    boot_denom_eps <- 0.3

    # Run analysis
    tryCatch({
      report <- simulate_fragment_ion_report(params, seed = 13)
      report <- compute_cov_unequal_replicates(report)
      resu <- run_ttests(report, boot_denom_eps = boot_denom_eps)

    }, error = function(e) {
      print("Retrying..")
      report <<- simulate_fragment_ion_report(params, seed = 10)
      report <- compute_cov_unequal_replicates(report)
      resu <<- run_ttests(report, boot_denom_eps = boot_denom_eps)

    })
    tables <- compute_contingency_tables(resu, alpha = 0.05)

    ## Note that params set only two conditions; so that there are only one
    ## comparison in each resulting 'tables' list
    all_nr_tables[[as.character(nr)]] <- tables[[1]]
  }

  NR_RESU[[i_s2]] <- all_nr_tables


  ## Plots ##
  pdf(paste0(SAVE_FOLDER, 
             substr(COMPARISON, 6, 9),
             "_meanstd", S1, "_betasum", BETA_SUM, "_",
             "plot_line_", substr(COMPARISON, 6, 9), "_",
             "bt", BETA_SUM, "dr", DIRI_SUM, "S1", S1, "s2", s2,
             ".pdf"))
  par(mar = c(5.1, 4.6, 4.1, 2.1))
  line_plot_contingency_tables(
    ALL_N_REPLICATES, all_nr_tables, scale_factor = 1 / 500, log = "x",
    ylab = "Sensitivity", ylim = c(0, 1),
    xlab = "Number of replicates", cex.lab = 1.7, cex = 2, lwd = 2.3,
    add_legend = TRUE, legend_coord = "bottomleft", legend_cex = 1.5)
  dev.off()
}

save(
  NR_RESU,
  file = paste0(SAVE_FOLDER,
                "nr_resu_", substr(COMPARISON, 6, 9), "_meanstd", S1,
                "_betasum", BETA_SUM, ".RData")
)
```