---
title: "MDstatsDIAMS for Precursor-Level Differential Analysis of DIA Data"
author: 
  - Namgil Lee^[Kangwon National University, and Bionsight Inc., namgil.lee@kangwon.ac.kr], Hojin Yoo^[Bionsight Inc.], Juhyoung Kim^[Kangwon National University], Heejung Yang^[Kangwon National University, and Bionsight Inc., heejyang@kangwon.ac.kr]
date: "June 23, 2025"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
---

This is a tutorial for performing downstream analysis with **MDstatsDIAMS** using
quantitative DIA reports obtained from various software platforms.

```{r import, message=FALSE}
library(MDstatsDIAMS)
```

## Converting DIA Reports into Standard Format

To convert various report formats into **MDstatsDIAMS** standard format,
**MDstatsDIAMS** provides both direct methods and indirect methods.
For direct methods, **Spectronaut**, **MaxQuant**, **Skyline**, and **MSstats**
report formats can be directly converted to standard report format.
For indirect methods, **MSstats** converters can convert various report formats
into **MSstats** report format, and **MDstatsDIAMS** can convert it into
standard report format. Here are some examples.

### Importing Spectronaut

A **Spectronaut** report can be converted into standard format directly.

```{r sn_to_standard, eval=FALSE}
sn_report <- arrow::read_parquet(
  paste0("/Users/namgil/Documents/Projects/MDstatsDIAMS/",
         "data/lip_quant_staurosporine_hela_sn_report.parquet")
)

standard_report <- convert_sn_to_standard(sn_report) 
```


### Importing MaxQuant

**MaxQuant** report consists `evidence.txt`, `msms.txt`, and `proteinGroups.txt`
files. A **MaxQuant** report can be converted a standard report as follows.

```{r mq_paths, eval=TRUE}
## Load MaxQuant evidence.txt and msms.txt files
mq_root <- paste0("/Users/namgil/Documents/Projects/MDstatsDIAMS/",
                  "data/lip_quant_staurosporine_hela_mq_report_4conds/")

mq_evidence_path <- paste0(mq_root, "evidence.txt")
mq_msms_path <- paste0(mq_root, "msms.txt")

mq_ev <- read.delim(mq_evidence_path)
mq_msms <- read.delim(mq_msms_path)

##  It is required to generate `annotation.txt' file in prior for **MSstats**.
mq_an <- create_annotation_df(
  evidence = mq_ev,
  n_replicates_per_condition = 4
)

## The **MaxQuant** is converted into standard format.
standard_report <- convert_mq_to_standard(mq_ev, mq_msms, mq_an)
```


### Importing Skyline

A **Skyline** transition report can be converted into standard format after an 
annotation data is prepared.

```{r skyline_to_msstats, eval=FALSE}
sk_report <- arrow::read_parquet(
  paste0("/Users/namgil/Documents/Projects/MDstatsDIAMS/",
         "data/lip_quant_staurosporine_hela_sk_transition_4conds.parquet")
)

# Make an annotation data frame
sk_an <- data.frame(
  Condition = rep(paste0("CON", c(1, 4, 5, 8)), each = 4),
  Replicate = paste0("StauroDoseResp-", c(1:4, 17:24, 33:36)),
  Run = paste0("StauroDoseResp-", c(1:4, 17:24, 33:36))
)

# Convert to standard format
standard_report <- convert_sk_to_standard(sk_report, annotation = sk_an)
```


## Running Statistical Tests of Mean Differences Between Two Groups

In **MDstatsDIAMS**, precursor-level differential analysis methods are 
available, including fundamental t-test methods (`paired`: paired t-test,
`independent`: independent samples t-test, `shrinkage`: shrinkage t-test),
as well as external methods (`msstatslip`: **MSstatsLiP**, `rots`: **ROTS**).

The `run_ttests()` will run statistical methods specified by `method_names =`
for comparing a base condition (e.g., "DMSO") with all the other conditions. If
`method_names = NULL` (default), all the available methods will be run.

```{r run_ttests, eval=FALSE}
## Select a subset of proteins to reduce time cost
n_proteins <- 100
set.seed(1111)
sampled <- sample(unique(standard_report$protein_id), n_proteins)

sample_report <- standard_report[standard_report$protein_id %in% sampled, ]

test_results <- run_ttests(
  report = sample_report, method_names = NULL
)

print(names(test_results))
print(names(test_results[[1]]))
```

The test results obtained by `run_ttests()` have `p.value` column in each
table for every comparisons. The **MDstatsDIAMS** can compute local false
discovery rate (lfdr) score and append it into the table as `lfdr` column.

```{r compute_lfdr, eval=FALSE}
lfdr_results <- compute_lfdr_result(test_results)
```


## Making Summary Tables and Plots

The **MDstatsDIAMS** can compute the numbers of precursors
that were found significant between two conditions by the table row
`Rejected = TRUE`. The `p.value` or an alternative significance score such
as `lfdr` can be supplied for determining 

```{r print_contingency, eval=FALSE}
tables <- compute_contingency_table(
  lfdr_results, alpha = 0.05, q_value_column = 'lfdr'
)

print(tables)
```

The **MDstatsMDIAMS** can draw line plots for the proportion of rejected
precursors across comparisons for every statistical methods.

```{r plot_line, eval=FALSE}
line_plot_contingency_tables(x = c(4, 5, 8), tables = tables, rejected = TRUE)
```

The **MDstatsMDIAMS** can draw bar plots, alternatively.

```{r plot_bar, eval=FALSE}
bar_plot_contingency_tables(tables = tables, rejected = TRUE)
```
