---
title: "LME for HFX-narrow"
author: "Namgil Lee"
date: "2022-08-15"
output: html_document
---

# LME for HFX-narrow

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(nlme)
library(lme4)
```


## Load Data

```{r load_data}
sn_report_dir <- "G:/Spectronaut_Reports/snuh202205_rapamycin_HELA_replicates/"
sn_report_fname <- "20220607_033909_RAPAMYCIN_HELA_HFX_narrow_Report.xls"

sn_report = read.delim(
  paste0(sn_report_dir, sn_report_fname), 
  header=TRUE
)

sn_report_unique = sn_report[!duplicated(sn_report[, c("R.Condition", "R.Replicate", "PG.ProteinAccessions", "EG.ModifiedSequence", "FG.Charge")]), ]

print(dim(sn_report_unique))

# # Load Piazza Hela Rapamycin Dose-Response Data
# sn_report_unique <- read.csv("unique_Report_20220321_180930_480-Exp18-RapaDR-PubSpecLib_Report.csv", header = TRUE)
#
# print(dim(sn_report_unique))
```

## Preprocessing

```{r prepro}
sn_report_prepro <- sn_report_unique %>% 
  mutate(FG.LogQuantity = log10(FG.MS2Quantity)) %>%
  filter(FG.LogQuantity >= 1)

print(dim(sn_report_prepro))
```

## Fit Linear Mixed Effect Models

```{r fix_type}
sn_report_prepro <- sn_report_prepro %>%
  mutate(R.Condition = factor(R.Condition),
         PG.ProteinAccessions = factor(PG.ProteinAccessions),
         PG.ProteinNames = factor(PG.ProteinNames),
         EG.PrecursorId = factor(EG.PrecursorId))
```

```{r prepare_proteins}
Proteins <- unique(sn_report_prepro$PG.ProteinAccessions)
PvaluesCondition <- rep(NA, length(Proteins))
PvaluesConditionPrecursor <- vector("list", length(Proteins))
Nprecursors <- rep(0, length(Proteins))
```

```{r sample_modeling}
count <- 2

data_protein_block <- sn_report_prepro %>%
    filter(PG.ProteinAccessions == Proteins[count])
num_precursor <- length(unique(data_protein_block$EG.PrecursorId))

print(paste("Test runs:"))

print(paste("Protein is", Proteins[count]))

print(paste("Number of precursors is", num_precursor))

fit_simple <- lme(
  FG.LogQuantity ~ R.Condition * EG.PrecursorId,
  random = ~ 1 + R.Condition | EG.PrecursorId,
  data = data_protein_block
)

print("Estimation result:")

print(fit_simple)

print("Summary for the estimation:")

print(summary(fit_simple))
```

```{r run_all_proteins, eval=FALSE}
for (count in 1:length(Proteins)) {

  print(paste("Running", count, "/", length(Proteins), "..."))

  data_protein_block <- sn_report_prepro %>%
    filter(PG.ProteinAccessions == Proteins[count])
  
  num_precursor <- length(unique(data_protein_block$EG.PrecursorId))

  if (num_precursor > 1) {
    fit_simple <- lme(
      FG.LogQuantity ~ R.Condition * EG.PrecursorId,
      random = ~ 1 + R.Condition | EG.PrecursorId,
      data = data_protein_block
    )
    PvaluesCondition[count] <- summary(fit_simple)$tTable[2, 5]
    PvaluesConditionPrecursor[[count]] <- summary(fit_simple)$tTable[(3+(num_precursor-1)):(2+2*(num_precursor-1)), 5]
  } else {
    fit_simple <- aov(
      FG.LogQuantity ~ R.Condition,
      data = data_protein_block
    )
  }
  PvaluesCondition[count] <- summary(fit_simple)[[1]]$"Pr(>F)"[1]
  PvaluesConditionPrecursor[[count]] <- NA
  Nprecursors[count] <- num_precursor
}

write.csv(data.frame(Proteins, Pvalues, Nprecursors), 
          file = "20220815-lme-HFX_narrow-results.csv", 
          row.names = FALSE)

