---
title: "LME for Staurosporine"
author: "Namgil Lee"
date: "2022-08-24"
output: html_document
---

# LME for Staurosporine

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(nlme)
library(lme4)
```


## Load Data

Spectronaut report data will be loaded from the following directory:

```{r load_data}
sn_report_dir <- "G:/Spectronaut_Reports/snuh2022/"
sn_report_fname <- "HELA_staurosporine_dose_202202_Report.xls"

sn_report = read.delim(
  paste0(sn_report_dir, sn_report_fname), 
  header=TRUE
)
```

Extract unique precursors from the original report:

```{r}
sn_report_unique = sn_report[!duplicated(sn_report[, c("R.Condition", "R.Replicate", "PG.ProteinAccessions", "EG.ModifiedSequence", "FG.Charge")]), ]

print(dim(sn_report_unique))

```

## Check Data Sizes 

### (1) Experimental conditions

```{r}
print(table(sn_report_unique$R.Condition))  # Dose

print(unique(sn_report_unique$R.Replicate))  # Repetition

print(paste("Number of raw files:", length(unique(sn_report_unique$R.FileName))))
```

### (2) Proteins and peptides


```{r}
length(unique(sn_report_unique$PG.ProteinAccessions))  # Protein

length(unique(sn_report_unique$PG.ProteinNames))

length(unique(sn_report_unique$PEP.GroupingKey))  # Modified Peptide

length(unique(sn_report_unique$EG.ModifiedSequence))

length(unique(sn_report_unique$PEP.StrippedSequence))  # Stripped Peptide (x)

length(unique(sn_report_unique$EG.PrecursorId))  # PrecursorID (Modified Peptide x charge)

```
```{r num_proteotypic}
count_protein_per_prec <- sn_report_unique %>% 
  group_by(EG.PrecursorId) %>% 
  summarise(count_protein = length(unique(PG.ProteinAccessions)))
print(paste("number of proteins for each precursor: ", 
            min(count_protein_per_prec$count_protein), 
            max(count_protein_per_prec$count_protein)))
```


## Explore Distribution of Y

The data is not separated for target v.s. non-target proteins.

Still, need to explore an 'extreme' skewness.

```{r load_known_target}
known_targets_df <- read.csv("G:/Spectronaut_Misc/KnownTargetList_Staurosporine.csv", header=TRUE)

known_targets <- known_targets_df$UniprotID
```


```{r prepro}
sn_report_prepro <- sn_report_unique %>% 
  mutate(FG.LogQuantity = log10(FG.MS2Quantity), 
         R.Condition = factor(
           R.Condition, 
           levels = c("DMSO", 2:9))
         ) 
```

```{r hist_quantity}
boxplot(FG.Quantity ~ R.Condition, data = sn_report_prepro, 
        main = "Precursor Quantity")
```
```{r hist_logquantity}
boxplot(FG.LogQuantity ~ R.Condition, data = sn_report_prepro, 
        main = "Precursor Quantity (Log)")
```
```{r hist_log_quantity}
quantity_counts = hist(sn_report_prepro$FG.LogQuantity, 100, 
                       main = "Precursor Quantity (Log)")
```

In order to check if the near-zero values are outliers or drug-effects, 
make histogram for each dose condition.

```{r hist_logquantity_dose}
par(mfrow = c(3,3))
for (cond in levels(sn_report_prepro$R.Condition)) {
  hist( subset(sn_report_prepro, R.Condition==cond)$FG.LogQuantity, 100,
        main = paste0("Precursor Quantity (Log),", cond))
}
```

It can be concluded that those small values may have been influenced by drug. 


```{r find_zero_precursor}
count_dmso_zerovalue <- sn_report_prepro %>% 
  filter(R.Condition == "DMSO") %>% 
  group_by(PG.ProteinNames, EG.PrecursorId) %>% 
  summarize(count_zero_value = sum(FG.LogQuantity < 1.7))
```


### ee





